ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-08-30 papers/GPUScA-2011/performance.tex

While we did not have an OpenMP code for multi-core comparisons, the transformed
OpenCL code on the NVIDIA C2050 was up to 40 times faster than the best serial
Fortran code executing on the host CPU.

>>>>

While we did not have an OpenMP code for multicore comparisons, the transformed
OpenCL code on the NVIDIA C2050 was up to 40 times faster than the best serial
Fortran code executing on the host CPU.

### 2011-08-30 papers/GPUScA-2011/programming-model.tex

A question that one may pose is ``\emph{Why choose Fortran and not a
  more modern language like X for programming accelerator
  architectures?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and
many-core accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.
However, for many scientific users with existing codes written in
Fortran, new languages and language extensions to use novel new
architectures present a challenge: how do programmers effectively use
them while avoiding rewriting code and potentially growing dependent
on a transient technology that will vanish tomorrow?  In this paper we
explore the constructs in Fortran that are particularly relevant to
GPU architectures.

@@

\noindent In this loop, it is necessary to manually implement a double buffering scheme
in order to avoid mixing values computed in the current loop with values from a 
prior execution of the loop. When shifts are employed, this can be expressed as:

@@

The static analysis and source-to-source transformations used in this work
require the programmer to use a language subset that employs a data-parallel
programming model.  In particular, it encourages the use of array notation,
elemental functions, and pure procedures.  From these language constructs, we
are able to easily transform Fortran procedures to a lower-level OpenCL kernel
implementation.

@@

It should be noted that in Fortran 2008,
the concept of impure elemental functions was introduced.  This requires that
elemental functions now need to be explicitly labelled as pure to ensure that they
are side effect free.

@@

It should be noted that the function {\tt region\_cpy} is pure and thus can be
called from within a pure kernel procedure, though it should be noted that {\tt
region\_ptr} is impure because it aliases the array parameter.  However as will
be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
introduce side effects in the functions that call it.  These two functions are
part of the language recognized by the compiler and though {\tt region\_cpy}
returns a copy of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics through
other means.  In addition to these two new functions, ForOpenCL provides the
compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
can be transformed to an OpenCL kernel and that the subroutine is pure
except for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in automated
identification of specific kernels to be transformed to OpenCL.  A directive-free
implementation would require the transformation tool be provided the set of 
kernels to work via a user defined list.

>>>>

A question that one may pose is ``\emph{Why choose Fortran and not a
  more modern language like X for programming accelerator
  architectures?}''  The recent rise in interest in concurrency and
parallelism at the language level due to multicore CPUs and
many-core accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.
However, for many scientific users with existing codes written in
Fortran, new languages and language extensions to use novel new
architectures present a challenge: how do programmers effectively use
them while avoiding rewriting code and potentially growing dependent
on a transient technology that will vanish tomorrow?  In this paper we
explore the constructs in Fortran that are particularly relevant to
GPU architectures.

@@

\noindent In this loop, it is necessary to manually implement a double buffering scheme
in order to avoid mixing values computed in the current execution of the loop with values from a 
prior execution of the loop. When shifts are employed, this can be expressed as:

@@

The static analysis and source-to-source transformations used in this work
require the programmer to use a language subset that employs a data-parallel
programming model.  In particular, it encourages the use of array notation,
pure elemental functions, and pure procedures.  From these language constructs, we
are able to easily transform Fortran procedures to a lower-level OpenCL kernel
implementation.

@@

It should be noted that in Fortran 2008,
the concept of impure elemental functions was introduced.  This requires that
elemental functions now need to be explicitly labeled as pure to indicate that they
are side effect free.

@@

It should be noted that the function {\tt region\_cpy} is pure and thus can be
called from within a pure kernel procedure, and {\tt
region\_ptr} is impure because it aliases the array parameter.  However as will
be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
introduce side effects in the functions that call it.  These two functions are
part of the language recognized by the compiler and though {\tt region\_cpy}
returns a copy of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics through
other means.  In addition to these two new functions, ForOpenCL provides the
compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
can be transformed to an OpenCL kernel and that the subroutine is pure
except for calls to {\tt region\_ptr}.  These directives are not strictly
necessary for the technique described in this paper, but aid in automated
identification of specific kernels to be transformed to OpenCL.  A directive-free
implementation would require the transformation tool be provided the set of 
kernels to work via a user defined list.
